Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2021-09-05 ms/ms.tex
UO changes
IK comments

% Abstract
\begin{abstract}
\noindent
\textbf{Background.}
During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
Overall, these interventions seem to have reduced the spread of the pandemic.
We hypothesized that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared, and that these differences can lead to errors in the estimation of the impact of NPIs. \\
\textbf{Methods.}
SEIR models were fitted to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates. The impact of NPIs was estimated from the inferred model parameters. \\
\textbf{Results.}
We infer mostly late effective start of interventions. For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective start date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI).
In one case we infer an early effective start date: Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\%~CI).
Moreover, we find that the impact of NPIs can be under-estimated if assuming they start at their official date.\\
\textbf{Conclusions.}
Differences between the official and effective start of NPIs are likely. Neglecting such differences can lead to under-estimation of the impact of NPIs, which could cause decision-makers to escalate interventions and guidelines. 
\\\\
{\small
\textbf{Keywords}: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
}
\end{abstract}

@@

%%% Fig - Re
% generated with
% > source ./Re.sh
% > python Fig_Re.py
\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
    \caption{
    \textbf{Impact of NPIs is under-estimated when assuming they start at their official date.}
    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on the epidemic.
     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$). The x-axis shows estimates when inferring the effective start of NPIs from the data (free~$\tau$). The dashed line shows a one-to-one correspondence.
     Markers and bars denote the posterior median and 50\% credible intervals (HDI).
    The relative reductions in $R$ are lower for the fixed~$\tau$ model (below the dashed line) In six of 12 regions.
    \autoref{fig:Re2} shows how the delay in effective start affects the estimation error. 
    AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
    } 
    \label{fig:Re}
\end{figure}

@@

\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig_Re2.pdf}
    \caption{
    \textbf{Under-estimation of impact of NPIs increases with the delay in their effective start.} The x-axis shows the days between the estimated effective start of NPIs, $\hat\tau$, and the official date, $\tau^*$. The y-axis shows the error in estimation of the impact of NPIs when assuming they start at their official date, i.e. the difference between the median estimate from a model with fixed~$\tau$ minus the estimate from a model with free~$\tau$, the y-axis and x-axis of \autoref{fig:Re}, respectively.
    Impact of NPIs defined as $\frac{R_1 - R_2}{R_1}$, where $R_1$ and $R_2$ are the effective reproduction numbers before and after the NPI, respectively (\autoref{eq:Re}).
	The dashed gray and black lines show a linear regression with and without SE, DK and CN, respectively
	(with: slope=$-0.019\pm0.026$~95\%~CI, intercept=$-0.105 \pm0.175$~95\%~CI, $R^2=0.211$;
     without: slope=$-0.016 \pm0.006$~95\%~CI, intercept=$-0.001 \pm0.036$~95\%~CI, $R^2=0.845$.)
	AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
    } 
    \label{fig:Re2}
\end{figure}

>>>>

% Abstract
\begin{abstract}
\noindent
\textbf{Background.}
During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
Overall, these interventions seem to have reduced the spread of the pandemic.
We hypothesized that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared, and that these differences can lead to errors in the estimation of the impact of NPIs. \\
\textbf{Methods.}
SEIR models were fitted to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates. The impact of NPIs was estimated from the inferred model parameters. \\
\textbf{Results.}
We infer mostly late effective start dates of interventions. For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective start date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI).
In one case we infer an early effective start date: Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\%~CI).
Moreover, we find that the impact of NPIs can be under-estimated if assuming they start at their official date.\\
\textbf{Conclusions.}
Differences between the official and effective start of NPIs are likely. Neglecting such differences can lead to under-estimation of the impact of NPIs, which could cause decision-makers to escalate interventions and guidelines. 
\\\\
{\small
\textbf{Keywords}: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
}
\end{abstract}

@@

%%% Fig - Re
% generated with
% > source ./Re.sh
% > python Fig_Re.py
\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
    \caption{
    \textbf{Impact of NPIs is under-estimated when assuming they start at their official date.}
    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on the epidemic.
     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$). The x-axis shows estimates when inferring the effective start of NPIs from the data (free~$\tau$). The dashed line shows a one-to-one correspondence.
     Markers and bars denote the posterior median and 50\% credible intervals (HDI).
    The relative reductions in $R$ are lower for the fixed~$\tau$ model (below the dashed line) in at least seven of 12 regions.
    Note that the estimated effective start date is uncertain for Sweden, Denmark, and Norway (\autoref{fig:tau-summary}).
    The effect of the delay in effective start affects on the estimation error is shown in \autoref{fig:Re2}.
    AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
    } 
    \label{fig:Re}
\end{figure}

@@

\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig_Re2.pdf}
    \caption{
    \textbf{Under-estimation of impact of NPIs increases with the delay in their effective start.} The x-axis shows the days between the estimated effective start of NPIs, $\hat\tau$, and the official date, $\tau^*$. The y-axis shows the error in estimation of the impact of NPIs when assuming they start at their official date, i.e. the difference between the median estimate from a model with fixed~$\tau$ minus the estimate from a model with free~$\tau$, the y-axis and x-axis of \autoref{fig:Re}, respectively.
    Impact of NPIs defined as $\frac{R_1 - R_2}{R_1}$, where $R_1$ and $R_2$ are the effective reproduction numbers before and after the NPI, respectively (\autoref{eq:Re}).
	The dashed gray and black lines show a linear regression with and without SE, DK and CN, respectively
	(with: slope=$-0.019\pm0.026$~95\%~CI, intercept=$-0.105 \pm0.175$~95\%~CI, $R^2=0.211$;
     without: slope=$-0.016 \pm0.006$~95\%~CI, intercept=$-0.001 \pm0.036$~95\%~CI, $R^2=0.845$.)
     Note that the estimated effective start date is uncertain for Sweden, Denmark, and Norway (\autoref{fig:tau-summary}).
	AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
    } 
    \label{fig:Re2}
\end{figure}
